BOAST: a Metaprogramming Framework to Produce Portable and Efficient Computing Kernels for HPC Applications

### 2015-03-31 Paper/BOAST.tex
Paper updated with D5.5 modifications.

\newcommand{\productname}[1]{\textsc{#1}}
\newcommand{\latin}[1]{\textit{#1}}

\newcommand{\Specfem}{\productname{Specfem 3D}\xspace}
\newcommand{\Cuda}{\productname{Cuda}\xspace}
\newcommand{\eg}{\latin{e.g.,}\xspace}
\newcommand{\OCL}{\productname{OpenCL}\xspace}
\newcommand{\code}[1]{\texttt{#1}}

\newcommand{\etc}[1]{\latin{etc}}

%\usepackage{lineno}
%\modulolinenumbers[5]
%\linenumbers

@@

@@

Porting and tuning HPC applications to new platforms is tedious and costly in
term of human resources.

Portability efforts are often lost when migrating to a new architecture, or code
lose maintainability because several versions of the code coexist, usually with
a lot of duplication.

@@

Genericity of HPC codes is often limited. Producing generic code in FORTRAN
90/95 is difficult as the language is not really  fit for it.

@@

@@

Evolution of HPC Architectures is rapid and also diverse: in the last 5
years no less than 6 architectures have been number one in the Top500:
\begin{itemize}
\item Intel Processor + Xeon Phi (Tianhe-2)
\item AMD Processor + NVIDIA GPU (Titan)
\item IBM BlueGene/Q (Sequoia)
\item Fujitsu SPARC64 (K computer)
\item Intel Processor + NVIDIA GPU (Tianhe-1)
\item AMD Processor (Jaguar)
\end{itemize}
Being able to efficiently use those architectures on such a small
time-frame is challenging. Stress frequency of architecture changes.

The race to exascale is not going to simplify the environment. All of the above
architectures can be considered. Network architectures are also part of the
architecture and can be very diverse.  For instance European FP7 project DEEP
considers using Accelerators (XEON Phi) while the European FP7 project
Mont-Blanc considers using low-power embedded processor with integrated GPU.

Running existing applications on those new architectures is an open project.
Thus, those projects have work packages dedicated to applications. Those
workpackages are dedicated to porting and optimizing Scientific applications on
those new architectures. In the DEEP project six applications were selected for
porting and optimizing, eleven were selected in the Mont-Blanc project.

Developed by physicist, chemist, meteorologist. Usually in FORTRAN for
historical and performance reason. Codes can be quite huge (several
thousands lines of code) with lots of functionalities. Nonetheless they are
usually based on computing kernels. Computing kernels are resource
intensive and well defined parts of a program with a limited set of input
and output. Those kernels are the time consuming part of an HPC
application. They are thus the prime target for optimization.

Those applications are often developed by several individuals. Sometimes some of
those developers only work a few month on the application. Maintaining optimized
code written by someone else is quite a challenge.

In Section~\ref{use_cases} we will present two HPC application that we used
as use cases: SPECFEM3D and BigDFT. They are both based on computing
kernels and were selected as candidate applications in the Mont-Blanc project.

The problematic here is to obtain computing kernels that present good
performances on the architectures encountered by the application while still
being portable after the optimization process took place. Indeed the application
might encounter one of the other architectures presented before. Investing
manpower to optimize the application for a new architecture is reasonable,
suffering hindrance from previous optimization work is not. Thus optimizations
have to be as orthogonal as possible from one another so as to be easily
activated and deactivated.

What we want is computing kernels that are:
\begin{itemize}
\item Written in a \emph{portable} manner,
\item Written in a way that raise developer \emph{productivity},
\item Written to present good \emph{performance}.
\end{itemize}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{BOAST_Workflow.pdf}
\caption{Structure and Workflow of the BOAST framework.}
\label{fig:boast_workflow}
\end{center}
\end{figure}

  The idea behind BOAST is to provide scientific application developers with a
framework to develop and test application computing kernels. The envisioned
workflow and program structure is presented in Figure~\ref{fig:boast_workflow}.
The user should start from an application kernel (implemented or just designed)
and implement it in a dedicated language (step 1). The language provides enough
flexibility for the kernel to be meta-programmed with several orthogonal
optimizations. Then the developer selects one configuration of the kernel and a
target language. Those parameter define the output source code that will be
generated by BOAST (step 2). The resulting code source is then built according to
the user specified compiler and options (step 3). If input data are available
then the kernel can be benchmarked and tested for non regression. Based on the
results, other optimizations can be selected (step 4) or new optimizations can
be added to the BOAST sources. The process can be repeated until a good
candidate is found on the target platform. The resulting kernel is then added to
the program (step 5).

Several of the steps in the workflow can be automated: optimization and compiler
flags exploration, non regression testing, benchmarking... This automation can
be done by scripting by the user or by interfacing other dedicated tools. In
order to achieve those goals three aspects should be considered:
\begin{itemize}
\item code description,
\item code generation,
\item and kernel execution runtime.
\end{itemize}

  \cite{videau2013boast}

@@

  \begin{itemize}
  \item Abstraction of programming concepts: variables, procedures (
portability, productivity )
  \item Operator overloading: simpler syntax ( productivity, portability )
  \item Two languages in one ( productivity, performance ) 
  \end{itemize}

Usually computing kernels are hotspots of an HPC application, and are most of
the time based on a loop nest. A lot of effort is dedicated to their tuning and
the obtained result is often quite different than the original procedure.
Several transformations can be applied to such kernels. And those optimization
are often applied manually as compiler may fail to recognise the opportunity.

@@

BOAST kernel description language should be able to express all these
optimizations. This gives us a set of constraints to implement in the language:
\begin{itemize}
\item Arbitrary number of variables have to be created and manipulated (types,
attributes...).
\item Procedures have to be abstracted (reunite FORTRAN and C like languages,
attributes...).
\item Functions must be available.
\item Variables, constants and functions must be composed in complex
Expressions.
\item Basic control structures (for, while, if/else...) have to be abstracted.
\item Powerful array management (several dimensions, transformations,
indexing...).
\end{itemize}

\begin{figure}
\begin{center}
\includegraphics[scale=0.8]{BOAST_DSL.pdf}
\caption{BOAST DSL intent.}
\label{fig:dsl}
\end{center}
\end{figure}

In order to manipulate those abstractions we want to have a syntax resembling
what programmers use. For instance commonly used operator have to be available
and behave as expected. It must also be possible to differentiate an action on
an abstraction in the context of an expression and in the context of the
management of this expression. This is why an embedded domain specific language
approach was selected~\cite{hudak1996building}. This allows for the coexistence
of two languages: the host language and the DSL. In our case the DSL allows the
description of the kernel while the host language provides the meta programming
of the kernel, cf Figure~\ref{fig:dsl}. Operator overloading of the host
language will provide the familiar syntax programmers are accustomed to.

It was also important to have our constructs like \emph{for} loops to have a
syntax approaching those commonly found in programming languages. For this we
needed a language which could seamlessly pass a block of code to a function.
Ruby~\cite{matsumoto2002ruby} is one such language. It has deep introspection
capabilities as well. This is the main reason it was selected.

    \subsubsection{BOAST keywords}

In order to clearly differentiate what is going to be generated from what is
related to manipulations in the host language four keywords were defined. They
are \emph{pr}, \emph{decl}, \emph{opn} and \emph{close}. As the language is
an EDSL, these four keywords are methods in the BOAST namespace.

@@

\begin{lstlisting}
int32_t i;
int64_t k;
double l[27];
i = 5;
k = (i + 5) * (2);
l[k - (-5)] = 1.0;
l[k + 1 - (-5)] = sin(i + 5);
\end{lstlisting}
\end{minipage}
\caption{BOAST Code Snippet for Variables and Expressions}
\label{fig:BOAST_algebra}
\end{figure}
The first and most fundamental abstraction is named Variable. Variables have a
type, a name and a set of named attributes. The existing attributes are mainly
inspired from FORTRAN. Those attributes are not limited and can be arbitrarily
enriched, allowing a lot of flexibility in Variable management.

The second abstraction is named Expression. It is used to combine variables into
algebraic or logic expressions. Most of the classical operators are overloaded
for those two abstractions and thus the syntax of the expressions are rather
straightforward. The exception is the assignment operator as it is important to
differentiate between assigning an Expression or a Variable in the ruby context
and the assignment operator in the context of a BOAST Expression. Thus the
assignment in a BOAST Expression is represented as the \emph{===} operator,
while the classical assignment is kept as the \emph{=} operator. Function calls
(FuncCall) are also abstracted and can be used in Expressions.

Figure~\ref{fig:BOAST_algebra} shows some basic usage of both Variables and
Expressions as well as the \emph{pr} and \emph{decl} keywords. For clarity we
stayed out of BOAST namespace so BOAST related class and methods are prefixed
with \emph{BOAST::}. Listing~\ref{lst:BOAST_algebra} shows the BOAST code that
when run produces the FORTRAN (Listing~\ref{lst:BOAST_algebra_FORTRAN}) and C
(Listing~\ref{lst:BOAST_algebra_C}) output. At first we define 2 Variables
\emph{i} and \emph{k} (note that \emph{k} is $64$ bit integer). The third variable
named \emph{l} is a one dimensional local array of a length of $27$, it is indexed
in the range $-5$ to $21$. All those Variables are affected to ruby variables
of the corresponding name.

On Line~4 we declare those three variables. Variable \emph{i} is then affected
the value 5. On Line~6 the \emph{j} ruby variable is used to store the BOAST
expression \emph{$i+5$}. This variable will the be used transparently through
the rest of the program.

On Line~8 we use the \emph{k} variable to index into the array \emph{l} using
the bracket operator. Would the array be multidimensional the index would be
coma separated, similar to FORTRAN notation.

@@

The classical control structures are implemented. \emph{If}, \emph{For},
\emph{While}, \emph{Case} are abstractions in BOAST matching the behavior of
corresponding control structures in other languages. An exception is the For in
BOAST more closely matches the for in FORTRAN than the one in C.

Figure~\ref{fig:BOAST_control} show some basic usage of the control structures.
The example a C macro or function that behaves similarly to the FORTRAN modulo
intrinsic. The sample script (inefficiently) computes and stores in \emph{j} the
number of multiples of $7$ in the 0 to 100 range. It uses the For and If control
structures. A Ruby block is passed to each of those constructs. This block is
evaluated at the time the construct is printed. If several such constructs are
needed (in an \emph{if elsif else} case for instance) they can be explicitly
passed as parameters using the \emph{lambda} Ruby keyword.

@@

\begin{lstlisting}
void magicfilter(const int32_t n, const int32_t ndat,
                 const double * x, double * y)
}
\end{lstlisting}
\caption{BOAST Code Snippet for Procedure}
\label{fig:BOAST_procedure}
\end{figure}

The last control structure is Procedure. It is used to describe either procedure
or functions. Figure~\ref{fig:BOAST_procedure} presents the use of this
abstraction. It is the signature of a real kernel from BigDFT. This kernel uses
an input array \emph{x} and an output array \emph{y} of double precision
numbers. Those array have two dimensions which depend on input variables
\emph{n} and \emph{ndat}. We can see here the use of the \emph{opn} and
\emph{close} keywords that are used to print a control structure without an
associated Ruby block. This time we placed ourselves inside of BOAST namespace.

The generated outputs in FORTRAN~\ref{lst:BOAST_procedure_FORTRAN} and
C~\ref{lst:BOAST_procedure_C} show the difference in meta-information that is
kept between both versions.

@@

\subsection{BOAST Runtime}

In the previous Section we presented BOAST's language. This allows us to
describe procedures and functions and to meta-program them using Ruby. But in
order to find the best version of a computing kernel each version has to be
compiled, linked and then executed to assess it's performance. This can be very
time consuming if this process cannot be automated. By enabling more versions to
be evaluated the automation will bring improved portability, better performance
and in the end improve the productivity of the developer.

In this section we will present the different aspects of BOAST runtime that
allow this automation. Those aspects are listed here:
  \begin{itemize}
  \item multi target language generation ( performance, portability )
  \item compilation ( productivity, performance )
  \item execution ( productivity, performance )
  \item tracer, dumper and replay for non regression tests ( productivity )
  \end{itemize}

  \subsubsection{Multi-target Language Generation}
\label{sec:multitarget}
Language availability and performance can vary between platforms. It is thus
important to be able to express computing kernels in different languages, based
on the availability and their respective merits on the target platform. Some
languages can have additional features, such as languages that target GPUs
(OpenCL, CUDA). The developer should be given tools to determine what kind of
language is currently used in order to be able to use those additional features.

Similarly the target language must be changed with ease in order to compare
different alternatives. Two methods are dedicated to this task: \emph{set\_lang}
and \emph{get\_lang}. The target language can also be set through environment
variable before launching BOAST, allowing for easy command line scripting.

@@

Compilation of the generated kernels must also be very flexible. Indeed, HPC
applications may encounter platforms with very diverse compilation environment.
Proprietary and dedicated compilers are common on HPC infrastructures. Thus
BOAST build system shows similar behavior than common build system. Compilers
and their compile/build options can be specified. This can be done at several
places. Here is the list by increasing order of precedence: BOAST configuration
file, environment variables and at kernel build time.

>>>>

\newcommand{\productname}[1]{\textsc{#1}}
\newcommand{\latin}[1]{\textit{#1}}
\newcommand{\eg}{\latin{e.g.,}\xspace}
\newcommand{\etc}[1]{\latin{etc}}

\newcommand{\Specfem}{\productname{Specfem 3D}\xspace}
\newcommand{\Cuda}{\productname{Cuda}\xspace}
\newcommand{\OCL}{\productname{OpenCL}\xspace}
\newcommand{\code}[1]{\texttt{#1}}

@@

\lstdefinestyle{customgputrace}{
  language=C,
  basicstyle=\scriptsize,
  backgroundcolor=\color{white},
  frame=single,
  captionpos=b,
  morekeywords={New, kernel, buffer, Mb, READ_WRITE, Buffer, written,
    b, at, out}
}

\lstdefinestyle{customcl}{
  language=C,
  basicstyle=\scriptsize,
  backgroundcolor=\color{white},
  frame=single,
  captionpos=b,
  morekeywords={__global, __local, global, local, __kernel, kernel, short,
uchar, uchar16, int16, short16}
}

@@

Porting and tuning HPC applications to new platforms is tedious and costly in
terms of human resources. Nonetheless, it is a very important aspect of the
Mont-Blanc project. Indeed, for the project, more than ten applications were
selected to be ported and optimized for the prototype platform.

Unfortunately, portability efforts are often lost when migrating to a new
architecture. Worse, code may lose maintainability because several versions of
some functionalities coexist, usually with a lot of duplication.

@@

Genericity of HPC codes is often limited. One of the reason is that producing
generic code in Fortran 90/95 is difficult as the language does not really fit
for it. Sometimes, adding genericity degrades performance as optimization
opportunities that come from over-specification are lost.

@@

Scientific Computing Applications are usually developed by physicist, chemist or
meteorologist. Those codes are usually written in Fortran for historical and
performance reasons. Codes can be quite huge (several thousands lines of code [LOC])
with lots of functionalities. Nonetheless, they are usually based on computing
kernels.  Computing kernels are resource intensive and well-defined parts of a
program that usually work on precisely defined data. Those kernels represent the most
time-consuming part of an HPC application, and consequently, they are the prime target for
optimization.

Those applications are often developed by several individuals. Sometimes, some
of those developers only work a few months on the application. Maintaining
optimized code written by someone else is quite a challenge. Several languages
and programming paradigms can also be used in a project and thus the maintainer
must be knowledgeable in several areas of expertise. Portability problems can
also be caused by the availability of an optimizing compiler for a specific
language and architecture. C compilers are usually the first available while
Fortran may sometimes arrive later.

In Section~\ref{use_cases} we will present two HPC applications that we used
as use cases: SPECFEM3D and BigDFT. They are both based on computing
kernels and were selected as candidate applications in the Mont-Blanc project.

The problematic here is to obtain computing kernels that present good
performances on the architectures encountered by the application while still
being portable after the optimization process took place. Indeed the
application might encounter one of the many architectures that can be found in
HPC. Investing manpower to optimize the application for a new architecture is
reasonable, suffering hindrance from previous optimization work is not. Thus
optimizations have to be as orthogonal as possible from one another so as to be
easily activated and deactivated.

What we want is computing kernels that are written:
\begin{itemize}
\item in a \emph{portable} manner,
\item in a way that raises developer \emph{productivity},
\item and, present good \emph{performance}.
\end{itemize}

@@

Evolution of HPC Architectures is rapid and also diverse: in the last 5
years no less than 6 architectures have been number one in the Top500:
\begin{itemize}
\item Intel Processor + Xeon Phi (Tianhe-2)
\item AMD Processor + NVIDIA GPU (Titan)
\item IBM BlueGene/Q (Sequoia)
\item Fujitsu SPARC64 (K computer)
\item Intel Processor + NVIDIA GPU (Tianhe-1)
\item AMD Processor (Jaguar)
\end{itemize}
Being able to efficiently use those architectures on such a small
time-frame is challenging.

The race to exascale is not going to simplify the environment. All of the above
architectures can be considered. Network architectures also can be very
diverse.  For instance European FP7 project DEEP considers using Accelerators
(XEON Phi) while the European FP7 project Mont-Blanc considers using low-power
embedded processor with integrated GPU.

Running existing applications on those new architectures is an open research
subject as well as an ongoing porting effort.  Thus, those projects have work
packages dedicated to applications. Those work packages are dedicated to porting
and optimizing Scientific applications on those new architectures. In the DEEP
project six applications were selected for porting and optimizing, eleven were
selected in the Mont-Blanc project.

\section{Motivating Example: OpenCL Laplace}
\label{sec:laplace}

During one of the Mont-Blanc face to face meeting a talk on OpenCL optimization
on the Mali GPU was given. One of the case study was a Laplace filter on the
GPU~\cite{opencl_arm_training}.

This is a good example of an algorithm that is simple in its formulation but
can be complex to optimize because many different optimizations are available
and can be combined together. The correct combination of optimizations will
depend on the targeted architecture.

\subsection{The Laplace Filter}

Listing~\ref{lst:laplace_c_filter} shows a simple implementation of the Laplace
filter in C99. For clarity, boundary conditions have been omitted.

\lstset{style=customc,
        caption={Laplace C99 Filter},
        breaklines=true,
        breakautoindent=false,
        label={lst:laplace_c_filter}}
\begin{lstlisting}
void laplace(const int width,
             const int height,
             const unsigned char src[height][width][3],
                   unsigned char dst[height][width][3]){
  for (int j = 1; j < height-1; j++) {
    for (int i = 1; i < width-1; i++) {
      for (int c = 0; c < 3; c++) {
        int tmp = -src[j-1][i-1][c] -   src[j-1][i][c] - src[j-1][i+1][c]\
                 - src[j  ][i-1][c] + 9*src[j  ][i][c] - src[j  ][i+1][c]\
                 - src[j+1][i-1][c] -   src[j+1][i][c] - src[j+1][i+1][c];
        dst[j][i][c] = (tmp < 0 ? 0 : (tmp > 255 ? 255 : tmp));
      }
    }
  }
}
\end{lstlisting}

A naive implementation in OpenCL is proposed in
Listing~\ref{lst:laplace_opencl_filter}. Each work item processes one pixel of
the resulting image.

\lstset{style=customcl,
        caption={Laplace OpenCL Filter},
        breaklines=true,
        breakautoindent=false,
        label={lst:laplace_opencl_filter}}
\begin{lstlisting}
kernel laplace(const int width,
               const int height,
               global const uchar *src,
               global       uchar *dst){
  int i = get_global_id(0);
  int j = get_global_id(1);
  for (int c = 0; c < 3; c++) {
    int tmp = -src[3*width*(j-1) + 3*(i-1) + c]\
             - src[3*width*(j-1) + 3*(i  ) + c]\
             - src[3*width*(j-1) + 3*(i+1) + c]\
             - src[3*width*(j  ) + 3*(i-1) + c]\
           + 9*src[3*width*(j  ) + 3*(i  ) + c]\
             - src[3*width*(j  ) + 3*(i+1) + c]\
             - src[3*width*(j+1) + 3*(i-1) + c]\
             - src[3*width*(j+1) + 3*(i  ) + c]\
             - src[3*width*(j+1) + 3*(i+1) + c];
    dst[3*width*j + 3*i + c] = clamp(tmp, 0, 255);
  }
}
\end{lstlisting}

\subsection{Possible Optimizations of the OpenCL Version}

Several optimizations were proposed and successively applied to the previous
implementation.

\paragraph{Vectorization}
The first proposed optimization involves computing five pixels instead of one using the
vectors available to the Mali architecture. The vectors have to reside in memory to
ensure an efficient execution. Using vectors of 16 elements, 15 useful components
are simultaneously loaded and can be used in computation. This optimization
yields speedups between 1.5 and 6 depending on the image size.

\lstset{style=customcl,
        caption={Laplace OpenCL Filter Vectorized},
        breaklines=true,
        breakautoindent=false,
        label={lst:laplace_opencl_filter_vector}}
\begin{lstlisting}
kernel laplace(const int width,
               const int height,
               global const uchar *src,
               global       uchar *dst){
  int i = get_global_id(0);
  int j = get_global_id(1);

  uchar16 v11_ = vload16( 0, src + 3*width*(j-1) + 3*5*i - 3 );
  uchar16 v12_ = vload16( 0, src + 3*width*(j-1) + 3*5*i     );
  uchar16 v13_ = vload16( 0, src + 3*width*(j-1) + 3*5*i + 3 );
  uchar16 v21_ = vload16( 0, src + 3*width*(j  ) + 3*5*i - 3 );
  uchar16 v22_ = vload16( 0, src + 3*width*(j  ) + 3*5*i     );
  uchar16 v23_ = vload16( 0, src + 3*width*(j  ) + 3*5*i + 3 );
  uchar16 v31_ = vload16( 0, src + 3*width*(j+1) + 3*5*i - 3 );
  uchar16 v32_ = vload16( 0, src + 3*width*(j+1) + 3*5*i     );
  uchar16 v33_ = vload16( 0, src + 3*width*(j+1) + 3*5*i + 3 );

  int16 v11 = convert_int16(v11_);
  int16 v12 = convert_int16(v12_);
  int16 v13 = convert_int16(v13_);
  int16 v21 = convert_int16(v21_);
  int16 v22 = convert_int16(v22_);
  int16 v23 = convert_int16(v23_);
  int16 v31 = convert_int16(v31_);
  int16 v32 = convert_int16(v32_);
  int16 v33 = convert_int16(v33_);

  int16 res = v22 * (int)9 - v11 - v12 - v13 - v21 - v23 - v31 - v32 - v33;
        res = clamp(res, (int16)0, (int16)255);
  uchar res_ = convert_uchar16(res);

  vstore8(res_.s01234567, 0, dst + 3*width*j + 3*5*i);
  vstore8(res_.s89ab,     0, dst + 3*width*j + 3*5*i + 8);
  vstore8(res_.scd,       0, dst + 3*width*j + 3*5*i + 12);
  dst[3*width*j + 3*5*i + 14] = res_.se;
}
\end{lstlisting}

\paragraph{Synthesizing Loads} It is possible to reduce the number of loads
since the vectors are overlapping.
Listing~\ref{lst:laplace_opencl_filter_vector_synth} shows how the vectors are
loaded in this case. This optimization yields marginal improvements.

\lstset{style=customcl,
        caption={Laplace OpenCL Filter Vectorized with Synthesized Loads},
        breaklines=true,
        breakautoindent=false,
        label={lst:laplace_opencl_filter_vector_synth}}
\begin{lstlisting}
  uchar16 v11_ = vload16( 0, src + 3*width*(j-1) + 3*5*i - 3 );
  uchar16 v13_ = vload16( 0, src + 3*width*(j-1) + 3*5*i + 3 );
  uchar16 v12_ = uchar16( v11_.s3456789a, v13_.s56789abc );
  uchar16 v21_ = vload16( 0, src + 3*width*(j  ) + 3*5*i - 3 );
  uchar16 v23_ = vload16( 0, src + 3*width*(j  ) + 3*5*i + 3 );
  uchar16 v22_ = uchar16( v21_.s3456789a, v23_.s56789abc );
  uchar16 v31_ = vload16( 0, src + 3*width*(j+1) + 3*5*i - 3 );
  uchar16 v33_ = vload16( 0, src + 3*width*(j+1) + 3*5*i + 3 );
  uchar16 v32_ = uchar16( v31_.s3456789a, v33_.s56789abc );
\end{lstlisting}

\paragraph{Temporary Variables Size} Using \textit{int} to store intermediary
results is unnecessary. Listing~\ref{lst:laplace_opencl_filter_short} show how
the code is modified to use smaller types. This yields a speedup of 1.3 for most
image sizes.

\lstset{style=customcl,
        caption={Laplace OpenCL Filter Vectorized Using \textit{short}},
        breaklines=true,
        breakautoindent=false,
        label={lst:laplace_opencl_filter_short}}
\begin{lstlisting}
  short16 v11 = convert_short16(v11_);
  short16 v12 = convert_short16(v12_);
  short16 v13 = convert_short16(v13_);
  short16 v21 = convert_short16(v21_);
  short16 v22 = convert_short16(v22_);
  short16 v23 = convert_short16(v23_);
  short16 v31 = convert_short16(v31_);
  short16 v32 = convert_short16(v32_);
  short16 v33 = convert_short16(v33_);

  short16 res = v22 * (short)9 - v11 - v12 - v13 - v21 - v23 - v31 - v32 - v33;
          res = clamp(res, (short16)0, (short16)255);
\end{lstlisting}

\paragraph{More Optimizations} Several other optimizations can be attempted at
this point. For instance reducing or increasing the number of pixels each work
item is processing. They can yield improved performance in some cases, especially
when avoiding memory alignment problems. These optimizations will not be shown
here but can be found in~\cite{opencl_arm_training}.

\subsection{Improving the Methodology}

The process described in the previous Subsection is quite tedious and requires
some intimate knowledge of the target architecture. The results are impressive,
as speedups of almost 10 are observed compared to the naive OpenCL
implementation.

Nonetheless, the process is frustrating. The optimizations are never evaluated
independently from one another. Some were arbitrarily configured (the number of
pixels chosen for instance). Testing the different combinations of those
optimizations and the different parameters would prove very costly in developer
time and like all repetitive and tedious tasks, error prone. Thus every created
kernel would have to be thoroughly tested to ensure no error was done.

In the next Section we will present our answer to this problematic: BOAST.
BOAST is a framework dedicated to kernel description, optimization, regression testing
and autotuning.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{BOAST_Workflow.pdf}
\caption{Structure and Work-flow of the BOAST framework.}
\label{fig:boast_workflow}
\end{center}
\end{figure}

BOAST provides scientific application developers with a framework to develop and
test application computing kernels~\cite{videau2013boast}.
Figure~\ref{fig:boast_workflow} illustrates the envisioned work-flow and program
structure.
The user starts from an application kernel (either designed or implemented), and
write it in a dedicated language (step 1).
The language provides enough flexibility for
the kernel to be meta-programmed with several orthogonal optimizations. Then the
developer selects a kernel configuration and a target language. Those
parameters define the output source code that will be generated by BOAST (step
2). The resulting code source is then built according to the user specified
compiler and options (step 3). If input data are available then the kernel can
be benchmarked and tested for non regression. Based on the results, other
optimizations can be selected (step 4) or new optimizations can be added to the
BOAST sources. The process can be repeated until a good candidate is found on
the target platform. The resulting kernel is then added to the program (step 5).

Several steps in the work-flow can be automated: optimization and compiler flags
exploration, non regression testing, benchmarking... This automation can be
scripted by the user or by interfacing other dedicated tools. In order
to achieve those goals three aspects should be considered:
\begin{itemize}
\item code description,
\item code generation,
\item and kernel execution run-time.
\end{itemize}

@@

%  \begin{itemize}
%  \item Abstraction of programming concepts: variables, procedures (
%portability, productivity )
%  \item Operator overloading: simpler syntax ( productivity, portability )
%  \item Two languages in one ( productivity, performance ) 
%  \end{itemize}

Usually computing kernels are hotspots of an HPC application, and they are most
of the time based on a loop nest. A lot of efforts are dedicated to their tuning
and the obtained result is often quite different from the original procedure.
Several transformations can be applied to such kernels. And those optimizations
are often applied manually as compiler may fail to recognise the opportunity.

@@

BOAST kernel description language should be able to express all these
optimizations. This gives us a set of constraints to implement in the language:
\begin{itemize}
\item Arbitrary number of variables have to be created and manipulated (types,
attributes...).
\item Procedures have to be abstracted (reunite Fortran and C like languages,
attributes...).
\item Functions must be available.
\item Variables, constants and functions must be composed in complex
Expressions.
\item Basic control structures (for, while, if/else...) have to be abstracted.
\item Powerful array management (several dimensions, transformations,
indexing...).
\end{itemize}

%\begin{figure}
%\begin{center}
%\includegraphics[scale=0.8]{BOAST_DSL.pdf}
%\caption{BOAST DSL intent.}
%\label{fig:dsl}
%\end{center}
%\end{figure}

In order to manipulate those abstractions we want to have a syntax resembling
what programmers use. For instance, commonly used operators have to be available
and behave as expected. It must also be possible to differentiate an action on
an abstraction in the context of an expression and in the context of the
management of this expression. $c = a + b$, an expression that affects the
results of $a + b$ to $c$  is not equal to $c \leftarrow a + b$, which saves the
expression $a + b$ to a variable $c$. This is why an embedded domain specific
language approach was selected~\cite{hudak1996building}. This allows for the
coexistence of two languages: the host language and the Domain Specific Language (DSL).
In our case, DSL allows the description of the kernel ($c = a + b$) while the host language
provides the meta programming of the kernel ($c \leftarrow a + b$)%, cf Figure~\ref{fig:dsl}
. Operator overloading of the host language will provide the familiar syntax
programmers are accustomed to. The host language will also provide easy
interfacing with libraries we might need during the development of our framework.

It was also important to have our constructs like \emph{for} loops to have a
syntax approaching those commonly found in programming languages. To this end, we
needed a language which could seamlessly pass a block of code to a function.
Ruby~\cite{matsumoto2002ruby} is one such language. It has deep introspection
capabilities as well. This is the main reason why it was selected.

    \subsubsection{BOAST Keywords}

In order to clearly differentiate what is going to be generated from what is
related to manipulations in the host language four keywords were defined. They
are \emph{decl}, \emph{pr}, \emph{opn} and \emph{close}. As the language is an
EDSL, these four keywords are methods in the BOAST namespace. Sample usage of
these keywords will be found in the next Figures.
%Usage of these keywords can be found on Figure~\ref{fig:keywords}.

@@

\begin{lstlisting}
int32_t i;
int64_t k;
double l[27];
i = 5;
k = (i + 5) * (2);
l[k - (-5)] = 1.0;
l[k + 1 - (-5)] = sin(i + 5);
\end{lstlisting}
\end{minipage}
\caption{BOAST Code Snippet for Variables and Expressions}
\label{fig:BOAST_algebra}
\end{figure}
The first and most fundamental abstraction is named Variable. Variables have a
type, a name and a set of named attributes. The existing attributes are mainly
inspired from Fortran. Those attributes are not limited and can be arbitrarily
enriched, allowing a lot of flexibility in Variable management.

The second abstraction is named Expression. It combines variables into
algebraic or logic expressions. Most of the classical operators are overloaded
for those two abstractions and thus the syntax of the expressions are rather
straightforward. The exception is the assignment operator as it is important to
differentiate between assigning an Expression or a Variable in the Ruby context
and the assignment operator in the context of a BOAST Expression. Thus the
assignment in a BOAST Expression is represented as the \emph{===} operator,
while the classical assignment is kept as the \emph{=} operator. Function calls
(FuncCall) are also abstracted and can be used in Expressions.

Figure~\ref{fig:BOAST_algebra} shows some basic usage of both Variables and
Expressions as well as the \emph{pr} and \emph{decl} keywords. For clarity we
stayed out of BOAST namespace so BOAST related class and methods are prefixed
with \emph{BOAST::}. Listing~\ref{lst:BOAST_algebra} shows the BOAST code that
produces the Fortran (Listing~\ref{lst:BOAST_algebra_FORTRAN}) and C
(Listing~\ref{lst:BOAST_algebra_C}) output. At first we define 2 Variables
\emph{i} and \emph{k} (note that \emph{k} is $64$ bit integer). The third
variable named \emph{l} is a one dimensional local array of length $27$, it
is indexed in the range $-5$ to $21$. All those Variables are affected to Ruby
variables of the corresponding name.

On Line~4 we declare those three variables. Variable \emph{i} is then affected
the value 5. On Line~6 the \emph{j} Ruby variable is used to store the BOAST
expression \emph{$i+5$}. This variable will the be used transparently through
the rest of the program.

On Line~8 we use the \emph{k} variable to index into the array \emph{l} using
the bracket operator. Would the array be multidimensional the index would be
comma separated, similar to Fortran notation.

@@

The classical control structures are implemented. \emph{If}, \emph{For},
\emph{While}, \emph{Case} are abstractions in BOAST matching the behavior of
corresponding control structures in other languages. An exception is the For in
BOAST more closely matches the for in Fortran than the one in C.

Figure~\ref{fig:BOAST_control} shows some basic usages of the control structures.
The example shows a C macro or function that behaves similarly to the Fortran modulo
intrinsic. The sample script (inefficiently) computes and stores in \emph{j} the
number of multiples of $7$ in the 0 to 100 range. It uses the For and If control
structures. A Ruby block is passed to each of those constructs. This block is
evaluated at the time the construct is printed. If several such constructs are
needed (in an \emph{if elsif else} case, for instance) they can be explicitly
passed as parameters using the \emph{lambda} Ruby keyword.

@@

\begin{lstlisting}
void magicfilter(const int32_t n, const int32_t ndat,
                 const double * x, double * y) {
}
\end{lstlisting}
\caption{BOAST Code Snippet for Procedure}
\label{fig:BOAST_procedure}
\end{figure}

The last control structure is Procedure. It describes either procedures
or functions. Figure~\ref{fig:BOAST_procedure} presents the use of this
abstraction. It illustrates the signature of a real kernel from BigDFT. This kernel uses
an input array \emph{x} and an output array \emph{y} of double precision
numbers. Those arrays have two dimensions which depend on input variables
\emph{n} and \emph{ndat}. We can see here the use of the \emph{opn} and
\emph{close} keywords that are used to print a control structure without an
associated Ruby block. This time we placed ourselves inside of BOAST namespace.

The generated outputs in Fortran (Listing~\ref{lst:BOAST_procedure_FORTRAN}) and
C (Listing~\ref{lst:BOAST_procedure_C}) show the difference in meta-information that is
kept between both versions.

@@

\subsection{BOAST Run-time}

In the previous Section we presented BOAST's language. This allows us to
describe procedures and functions and to meta-program them using Ruby.
Each version has to be compiled, linked and executed to assess its performance
in order to find the best version of a computing kernel.  This can be very
time consuming if this process cannot be automated. By enabling more versions
for evaluation, the automation will bring improved portability, better performance,
and in the end will improve the productivity of the developer.

In this section we will present the different aspects of BOAST run-time that
allow this automation. Those aspects include:
  \begin{itemize}
  \item multi target language generation ( performance, portability )
  \item compilation ( productivity, performance )
  \item execution ( productivity, performance )
  \item tracer, dumper and replay for non regression tests ( productivity )
  \end{itemize}

  \subsubsection{Multi-target Language Generation}
\label{sec:multitarget}
Language availability and performance varies between platforms. It is thus
important to express computing kernels in different languages, based
on the availability and their respective merits on the target platform. Some
languages have additional features, such as languages that target GPUs
(OpenCL, CUDA). The developer should be given tools to determine what kind of
language is currently used in order to be able to use those additional features.

Similarly the target language must be changed with ease in order to compare
different alternatives. Two methods are dedicated to this task: \emph{set\_lang}
and \emph{get\_lang}. The target language can also be set through an environment
variable before launching BOAST, allowing for easy command line scripting.

@@

Compilation of the generated kernels must also be very flexible because HPC
application developers may encounter platforms with very diverse compilation environments.
Proprietary and dedicated compilers are common on HPC infrastructures. Thus
BOAST build system exhibits similar behavior to common build systems. Compilers
and their compile/build options can be specified at several
places. The list by increasing order of precedence includes: BOAST configuration
file, environment variables, and at kernel build time.
